---
title: "ISRL Ch5 Lab EMALCA 2025"
author: "Angel Pineda"
date: "2025-06-26"
output: html_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

# Cross-Validation

In this lab, we explore the resampling techniques covered in this chapter. Some of the commands in this lab may take a while to run on your computer.

## The Validation Set Approach

We explore the use of the validation set approach in order to estimate the test error rates that result from fitting various linear models on the `Auto` data set.

Before we begin, we use the `set.seed()` function in order to set a *seed* for `R`'s random number generator, so that the reader of this book will
obtain precisely the same results as those shown below. It is generally
a good idea to set a random seed when performing an analysis such as cross-validation that contains an
element of randomness, so that the results obtained can be reproduced precisely at a later time.

We begin by using the
`sample()` function to split the set of observations into two halves, by selecting a random subset of $196$ observations out of the original $392$ observations. We  refer
to these observations as the training set.

```{r chunk1}
library(ISLR2)
set.seed(1)
train <- sample(392, 196)
```

(Here we use a shortcut in the sample command; see `?sample` for details.)
We then use the `subset` option in `lm()` to fit a linear regression using only the observations corresponding to the training set.

```{r chunk2}
lm.fit <- lm(mpg ~ horsepower, data = Auto, subset = train)
```

We now  use
 the `predict()` function to estimate the response for all $392$ observations,  and
 we  use
  the `mean()` function to calculate the MSE of the $196$ observations in the validation set. Note that the `-train` index below selects  only the observations that are not in the training set.

```{r chunk3}
attach(Auto)
mean((mpg - predict(lm.fit, Auto))[-train]^2)
```

Therefore, the estimated test MSE for the linear regression fit is $23.27$. We can use the `poly()` function to estimate the test error for the quadratic and cubic regressions.

```{r chunk4}
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train]^2)
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit3, Auto))[-train]^2)
```

These error rates are $18.72$ and $18.79$, respectively.
If we choose a different training set instead, then we will obtain somewhat different errors on the validation set.

```{r chunk5}
set.seed(2)
train <- sample(392, 196)
lm.fit <- lm(mpg ~ horsepower, subset = train)
mean((mpg - predict(lm.fit, Auto))[-train]^2)
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train]^2)
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, 
    subset = train)
mean((mpg - predict(lm.fit3, Auto))[-train]^2)
```

Using this split of the observations into a training set and a validation set,
we find that the validation set error rates for the models with linear, quadratic, and cubic terms are $25.73$, $20.43$, and $20.39$, respectively.

These results are consistent with our previous findings: a model that predicts `mpg` using a quadratic function of `horsepower` performs better than a model that involves only a linear function of `horsepower`, and there is little evidence in favor of a model that uses a cubic function of `horsepower`.

## Leave-One-Out Cross-Validation

The LOOCV estimate can be automatically computed for any generalized linear model  using the `glm()` and `cv.glm()` functions.  In the lab for Chapter 4, we used the `glm()` function to perform logistic regression by passing in  the `family = "binomial"` argument.
 But if we use `glm()` to fit a model without passing in the `family` argument, then it  performs linear regression, just like the `lm()` function.
So for instance,

```{r chunk6}
glm.fit <- glm(mpg ~ horsepower, data = Auto)
coef(glm.fit)
```

and

```{r chunk7}
lm.fit <- lm(mpg ~ horsepower, data = Auto)
coef(lm.fit)
```

 yield identical linear regression models. In this lab, we will  perform linear regression using
 the `glm()` function rather than the `lm()` function because the former can be used together with
`cv.glm()`. The `cv.glm()` function is part of the `boot` library.

```{r chunk8}
library(boot)
glm.fit <- glm(mpg ~ horsepower, data = Auto)
cv.err <- cv.glm(Auto, glm.fit)
cv.err$delta
```

The `cv.glm()` function produces a list with several components.  The two numbers in the `delta` vector contain the cross-validation results. In this case the numbers are identical (up to two decimal places) and correspond to the LOOCV statistic given in (5.1). Below, we discuss a situation in which the two numbers differ. Our cross-validation estimate for the test error is approximately $24.23$.

We can repeat this procedure for increasingly complex polynomial fits.
 To automate the process, we use the  `for()` function to initiate a *for loop* which iteratively fits polynomial regressions for polynomials of order $i=1$ to $i=10$, computes the associated cross-validation error, and stores it in the $i$th element of the vector `cv.error`.
 We begin by initializing the vector. 

```{r chunk9}
cv.error <- rep(0, 10)
for (i in 1:10) {
  glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.error[i] <- cv.glm(Auto, glm.fit)$delta[1]
}
cv.error
```

As in Figure 5.4, we see a sharp drop in the estimated test MSE between the linear and quadratic fits, but then no clear improvement from using higher-order polynomials.

## $k$-Fold Cross-Validation

The `cv.glm()` function can also be used to implement $k$-fold CV. Below we use $k=10$, a common choice for $k$, on the `Auto` data set.
We once again set a random seed and initialize a vector in which we will store the CV errors corresponding to the polynomial fits of orders one to ten.

```{r chunk10}
set.seed(17)
cv.error.10 <- rep(0, 10)
for (i in 1:10) {
  glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.error.10[i] <- cv.glm(Auto, glm.fit, K = 10)$delta[1]
}
cv.error.10
```

Notice that the computation time is shorter than that of LOOCV.
(In principle, the computation time for LOOCV for a least squares linear model should be faster than for $k$-fold CV, due to the availability
of the formula (5.2) for LOOCV; however, unfortunately the `cv.glm()` function does not make use of this formula.)
We still see little evidence that using cubic or higher-order polynomial terms leads to lower test error than simply using a quadratic fit.

We saw in Section 5.3.2 that the two numbers associated with `delta` are essentially the same when LOOCV is performed.
When we instead perform $k$-fold CV, then the two numbers associated with `delta` differ slightly. The first is the standard $k$-fold CV estimate,
as in (5.3). The second is a bias-corrected version. On this data set, the two estimates are very similar to each other.


